---
title: "Tutorial for using Climate Engine Time-series API"
date: "May 31, 2023"
output: html_document
description: "Tutorial for using Climate Engine Time-series API"
---

```{r setup, message = FALSE}
library(tidyverse) # Data processing and visualization
library(httr) # HTTP API requests
library(httr2) # HTTP API requests
library(DT) # Formatting data tables
```

## Preliminary test

Define Climate Engine API key and run simple endpoint to get key expiration to ensure that the key authenticated correctly.

```{r test}
# Define root url for Climate Engine API
root_url <- 'https://api.climateengine.org/'

# Define key
key = 'key_here'

# Define endpoint
endpoint = '/home/key_expiration'

# Run simple endpoint to get key expiration
test_r <- GET(paste0(root_url, endpoint), config = add_headers(Authorization = key))
print(content(test_r, "text",  encoding = "ISO-8859-1"))
```

## Get time-series data

Get time-series data for the GridMET Long-term Drought Blend. 

Learn more about this API endpoint here: https://docs.climateengine.com/docs/build/html/timeseries.html#timeseries-native-points

Learn more about drought blends here: https://www.drought.gov/data-maps-tools/multi-indicator-drought-index-midi

```{r request-ltb}
# Define endpoint for initial data request
endpoint <- "/timeseries/native/points"

# Define API arguments time-series endpoint to get long-term blend data 
ltb_query <- list(dataset = 'GRIDMET_DROUGHT',
variable = "long_term_blend",
start_date = '2000-01-01',
end_date = '2022-12-31',
coordinates = '[[-121.98,39.03]]',
area_reducer = 'mean')

# Run GET request to get data
ltb_r <- GET(paste0(root_url, endpoint), config = add_headers(Authorization = key), query = ltb_query)
print(ltb_r)
```

A status code of 200 indicates that the data was returned successfully. In the next step, we'll unpack the JSON that was returned from the API and convert it to a data frame.

## Parse time-series data

Parse time-series of the long-term blend drought data from GridMET DROUGHT from a JSON to data frame

```{r parse-ltb}
# Parse JSON returned from API
ltb_r_parse <- content(ltb_r, "parsed")
ltb_df <- enframe(unlist(ltb_r_parse[[1]]$Data))
ltb_cols <- ltb_df$name %>% unique()

# Define simple function to iterate over "name" values in parsed dataframe and return single column dataframe with those values
generate_data_frame <- function(col, df){

  # Filter for name from parsed API return
  df_val <- df %>%
    filter(name == col)

  # Create output data frame based on name from parsed API return
  out_df <- tibble(val = df_val$value)

  # Provide names from parsed API return as column name
  colnames(out_df) <- c(col)

  return(out_df)
}

# Map generate_data_frame over list of columns and clean up resulting data frame
ltb_r_df <- map(ltb_cols, generate_data_frame, ltb_df) %>%
  bind_cols() %>%
  mutate(Date = as.Date(Date),
         Value = as.numeric(long_term_blend),
         Variable = "long_term_blend") %>%
  filter(Value > -10) %>%
  select(-long_term_blend)
datatable(ltb_r_df)
```


## Visualize time-series data
Make a simple plot of the time-series data

```{r visualize-ltb}
# Make a simple plot of the time-series data
ggplot(data = ltb_r_df)+
  geom_line(mapping = aes(x = Date, y = Value))+
  theme_minimal()
```


## Get and parse additional time-series data

Get time-series data for the GridMET Short-term Drought Blend. 

Learn more about this API endpoint here: https://docs.climateengine.com/docs/build/html/timeseries.html#timeseries-native-points

Learn more about drought blends here: https://www.drought.gov/data-maps-tools/multi-indicator-drought-index-midi

```{r request-stb}
# Define endpoint for initial data request
endpoint <- "/timeseries/native/points"

# Define API arguments time-series endpoint to get long-term blend data 
stb_query <- list(dataset = 'GRIDMET_DROUGHT',
variable = "short_term_blend",
start_date = '2000-01-01',
end_date = '2022-12-31',
coordinates = '[[-121.98,39.03]]',
area_reducer = 'mean')

# Run GET request to get data
stb_r <- GET(paste0(root_url, endpoint), config = add_headers(Authorization = key), query = stb_query)

# Parse JSON returned from API
stb_r_parse <- content(stb_r, "parsed")
stb_df <- enframe(unlist(stb_r_parse[[1]]$Data))
stb_cols <- stb_df$name %>% unique()

# Map generate_data_frame over list of columns and clean up resulting data frame
stb_r_df <- map(stb_cols, generate_data_frame, stb_df) %>%
  bind_cols() %>%
  mutate(Date = as.Date(Date),
         Value = as.numeric(short_term_blend),
         Variable = "short_term_blend") %>%
  filter(Value > -10)
datatable(stb_r_df)
```


## Visualize short-term and long-term blend time-series data

Combine long-term and short-term blend time-series to plot them on the same graph

```{r visualize-blends, exercise = TRUE, exercise.setup = "request-stb"}
# ----- Bind rows of stb and ltb data frames -----
stb_ltb_r_df <- bind_rows(stb_r_df, ltb_r_df) %>%
  mutate(Variable = ifelse(Variable == 'long_term_blend', 'Long-term blend', 'Short-term blend'))

# ----- Make a simple plot of the time-series data ----- 
ggplot(data = stb_ltb_r_df)+
  geom_line(mapping = aes(x = Date, y = Value, color = Variable), size = 0.3)+
  labs(title = "Long-term vs Short-term Drought Blends, GridMET Drought (2000-2023)") +
  theme_minimal()+
  theme(legend.position="bottom",
        axis.title.x = element_blank(),
        legend.title=element_blank())
```