---
title: "Tutorial for using OpenET Time-series Data in Climate Engine API"
date: "January 7, 2026"
output: html_document
description: "Tutorial for using OpenET Time-series Data in Climate Engine API"
---

```{r setup, message=FALSE}
library(tidyverse) # Data processing and visualization
library(httr) # HTTP API requests
library(httr2) # HTTP API requests
library(DT) # Formatting data tables
library(sf) # Importing shapefile
library(mapview) # Visualizing shapefile
```

## Preliminary test

Define Climate Engine API key and run simple endpoint to get key expiration to ensure that the key authenticated correctly:

```{r test}
# Define root url for Climate Engine API
root_url <- "https://api.climateengine.org/"

# Define key
key = 'key_here'

# Define endpoint
endpoint = '/home/key_expiration'

# Run simple endpoint to get key expiration
test_r <- GET(paste0(root_url, endpoint), config = add_headers(Authorization = key))
print(content(test_r, "text",  encoding = "ISO-8859-1"))

# Clean up unneeded objects
rm(test_r, endpoint)
```

## Import field boundaries

These are sample field boundaries, but you can run similar analysis for any shapefile. We will import the field boundaries as an sf object and visualize them on a leaflet map using the mapview package.

```{r}
# Import the field boundaries shapefile
fields <- st_read('./data/fields.shp')

# Visualize the fields on a leaflet map
mapview(fields)
```

## Select a field to get OpenET data for

Here, we will select an individual field to get OpenET data for and will parse a character string of the coordinates to pass to the API request in the next code chunk.

```{r select-field}
# Get list of field IDs
fields_ids <- fields$OBJECTID

# Select field to get data for
field <- fields %>% filter(OBJECTID == fields_ids[2])

# Function to parse coordinates for API request from sf polygon object (shapefile)
parse_coordinates <- function(polygon){
  
  # Helper function to parse coordinate pair from dataframe
  get_coord <- function(i){
    x <- coords_df[i, 1] %>% unlist()
    y <- coords_df[i, 2] %>% unlist()
    coord <- paste0('[', x, ',', y, ']')
    return(coord)
  }
  
  # Concatenate coordinate string from individual coords
  coords_df <- as_tibble(st_coordinates(polygon))
  coords_inner <- paste(map(seq(1, nrow(coords_df)), get_coord) %>% unlist(), collapse = ',')
  coords_str <- paste0('[[', coords_inner, ']]')
  
  return(coords_str)
}

field_coords <- parse_coordinates(field)
```

## Make API requests to get OpenET data

```{r request-openet}
# Define dataset for function below
dataset = "OPENET_CONUS"

# Make a list of ET Model options — we will loop over these using the function below to return all of the data
models <- c('et_ensemble_mad', 'et_eemetric','et_ssebop','et_sims','et_geesebal','et_ptjpl','et_disalexi')

# Function to return data frame of OpenET time-series
get_openet_df <- function(model){
  
  print(paste0("Requesting data for ", model))
  
  # Define Climate Engine API Point 
  endpoint <- 'timeseries/native/coordinates'
  
  # Create list of API parameters for request
  params = list(
    dataset = dataset,
    variable = model,
    start_date = '2016-01-01',
    end_date = '2021-12-01',
    coordinates = field_coords,
    area_reducer = 'mean')

  # Make API request
  model_r <- GET(paste0(root_url, endpoint), config = add_headers(Authorization = key), query = params)

  # Parse JSON returned from API
  model_r_parse <- content(model_r, "parsed")
  model_df <- enframe(unlist(model_r_parse[[2]][[1]]$Data))
  model_cols <- model_df$name %>% unique()

  # Function to iterate over "name" values in parsed dataframe and return single column dataframe with those values
  generate_data_frame <- function(col, df){
  
    # Filter for name from parsed API return
    df_val <- df %>%
      filter(name == col)
  
    # Create output data frame based on name from parsed API return
    out_df <- tibble(val = df_val$value)
  
    # Provide names from parsed API return as column name
    colnames(out_df) <- c(col)
  
    return(out_df)
  }

  # Map generate_data_frame over list of columns and clean up resulting data frame
  model_r_df <- map(model_cols, generate_data_frame, model_df) %>%
    bind_cols() %>%
    mutate(Date = as.Date(Date),
           Value = as.numeric(.[[2]]),
           Variable = paste0(model)) %>%
    filter(Value > -10) %>%
    select(-2)
  
  return(model_r_df)
}

# Map OpenET API function over list of models to get timeseries from all models
openet_df <- map(models, get_openet_df) %>% 
  bind_rows() # Combine list of dataframes into an individual dataframe
datatable(openet_df)
```

## Visualize the OpenET model data

Make simple plots of the OpenET time-series data

```{r visualize-openet}
# Make a simple plot of the time-series data
ggplot(data = openet_df)+
  geom_line(mapping = aes(x = Date, y = Value, color = Variable, size = Variable, alpha = Variable))+
  labs(title = paste0("OpenET models for field #", field$FieldID %>% unlist()), y = 'Monthly evapotranspiration (mm)') +
  scale_colour_manual(values = c('et_ensemble_mad' = 'black', 'et_eemetric' = "#F8766D", 'et_ssebop' = "#B79F00", 'et_sims' = "#00BA38", 'et_geesebal' = "#00BFC4", 'et_ptjpl' = "#619CFF", 'et_disalexi' = "#F564E3"), labels = c("Ensemble mean", "eeMETRIC", "SSEBop", "SIMS", "geeSEBAL", "PT-JPL", "DisALEXI"))+
  scale_size_manual(values = c('et_ensemble_mad' = 1, 'et_eemetric' = 0.5, 'et_ssebop' = 0.5, 'et_sims' = 0.5, 'et_geesebal' = 0.5, 'et_ptjpl' = 0.5, 'et_disalexi' = 0.5), labels = c("Ensemble mean", "eeMETRIC", "SSEBop", "SIMS", "geeSEBAL", "PT-JPL", "DisALEXI"))+
    scale_alpha_manual(values = c('et_ensemble_mad' = 1, 'et_eemetric' = 0.7, 'et_ssebop' = 0.7, 'et_sims' = 0.7, 'et_geesebal' = 0.7, 'et_ptjpl' = 0.7, 'et_disalexi' = 0.7), labels = c("Ensemble mean", "eeMETRIC", "SSEBop", "SIMS", "geeSEBAL", "PT-JPL", "DisALEXI"))+
  theme_minimal()+
  theme(legend.position="bottom",
        axis.title.x = element_blank(),
        legend.title=element_blank())
```
